1.  Session Set up :
```{r, message = FALSE}
library(tidyverse)
library(lubridate)  
library(zoo)
library(trend)
library(here)
library(ggplot2)
library(Kendall)
getwd()

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"),
        panel.background = element_rect(fill ="white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        plot.title = element_text(
          hjust = 0.5,         
          face = "bold"),
        legend.position = "middle")


```
2. Import the ten datasets from the time_series folder in the egrid_Data data folder. 
```{r, message = FALSE}

folder <- "~/DataExplorationFinalProjetc/egrid_Data/time_series"

files <- list.files(
  folder, 
  pattern = "\\.csv$",    # match .csv files
  full.names = TRUE
)

time_series <- files %>%
  set_names(basename(.)) %>%     # keeps track of source file
  map_dfr(~ read_csv(.x, show_col_types = FALSE), .id = "source_file")

glimpse(time_series)


```
3. Set date column as a date class.

4. Wrangle  dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame time_series 
```{r, message = FALSE}
time_series$Year <- as.Date(time_series$Datayear, format = "%m/%d/%Y")

time_series <- time_series[, c("Date", 
                                   "Daily.Max.8.hour.Ozone.Concentration", 
                                   "DAILY_AQI_VALUE")]
head(time_series)

# 5
Days <- as.data.frame(seq(from = as.Date("2010-01-01"),
                          to   = as.Date("2019-12-31"),
                          by   = "day"))
names(Days)[1] <- "Date"
head(Days)

# 6
time_series$Date <- as.Date(time_series$Date)
Days$Date <- as.Date(Days$Date)

# Now join again
time_series <- left_join(Days, time_series, by = "Date")

# Check
dim(time_series)
head(time_series)
```
